Topics in Statistical Theory --- Example Sheet 1

### 2020-10-18 modern_statistical_methods/msm_ES1.tex
updates topics summary and example sheets for msm & topics

To compute $\hat\beta$, we differentiate the objective function w.r.t.\ $\beta$: 
\begin{align*}
    \nabla_\beta \norm{Y- \mu \vec 1 - X\beta}_2^2 + \lambda \norm{\beta}_2^2 &= ...
\end{align*}
\TODO Finish
\end{solution}

@@

\end{document}

>>>>

The computation of $\hat\beta$ is given in the lecture notes. 
\end{solution}

@@

\begin{question}
	Prove Proposition 4 in the notes. \emph{Hint: for part (ii) it may help to consider the eigendecompositions of positive semi-definite matrices $K^{(1)}$ and $K^{(2)}$ derived from kernels $k_1$ and $k_2$ in the form $K^{(1)} = PDP\T = \sum_{i=1}^n P_iP_i\T D_{ii}$ for example. }
\end{question}

\begin{proof}
	Proposition 4 is the following:
	\begin{mdframed}
		Suppose $k_1, k_2, \dotsc$ are kernels. 
		\begin{enumerate}[(i)]
			\item If $\alpha_1, \alpha_2 \geq 0$ then $\alpha_1 k_1 + \alpha_2 k_2$ is a kernel. If $\lim_{k\to\infty} k(x, x') \eqqcolon k(x, x')$ exists for all $x, x' \in \Cal X$, then $k$ is a kernel. 
			\item The pointwise product $k = k_1k_2$ is a kernel.
		\end{enumerate}
	\end{mdframed}

Let $x_1, \dotsc, x_n \in \Cal X$ and define the positive definite matrix $K^{(m)}$ by $K^{(m)}_{i, j} = k_m(x_i, x_j)$. 
\begin{enumerate}[(i)]
	\item Since $(\alpha_1 k_1 + \alpha_2 k_2)(x_i, x_j) = \alpha_1 k_1(x_i, x_j) + \alpha_2 k_2(x_i, x_j)$, the kernel matrix $K_{ij} \ceq (\alpha_1 k_1 + \alpha_2 k_2)(x_i, x_j)$ equals $\alpha_1 K_1 + \alpha_2 K_2$, and is therefore clearly symmetric positive semi-definite. This shows that $\alpha_1 k_1 + \alpha_2 k_2$ is a kernel.

	For the second part (the pointwise limit of kernels is a kernel), it suffices to show that the entrywise limit of positive semi-definite matrices is again positive semidefinite. But this is clear: if $K^{(m)} \to K$ entrywise, then for any $\vec x \in \RR^n$ we have
	\[
	\vec x\T K \vec x = \vec x\T (\lim_{m\to\infty} K^{(m)}) \vec x = \lim_{m\to\infty} \vec x\T K^{(m)}\vec x \geq 0,
	\]
	so $K$ is positive semi-definite.

	\item Let $K^{(1)}$ and $K^{(2)}$ be positive semi-definite, then we must show that the pointwise product $K^{(1)} \circ K^{(2)}$ is positive semi-definite as well. Write $K^{(1)} = \sum_{\ell=1}^n \lambda_\ell P_\ell P_\ell\T$ and $K^{(2)} = \sum_{\ell=1}^n \mu_\ell Q_\ell Q_\ell\T$, where $\lambda_i, \mu_i \geq 0$ and $\qty{P_\ell}$, $\qty{Q_\ell}$ are orthonormal bases of $\RR^n$.

	Then we find
	\begin{align*}
	(K^{(1)} \circ K^{(2)})_{ij} &= \qty(\sum_m \lambda_m P_{im} P_{jm}) \qty(\sum_\ell \mu_\ell Q_{i\ell} Q_{j\ell}) = \sum_{m, \ell} \lambda_m \mu_\ell (P_{im} P_{jm} Q_{i\ell} Q_{j\ell}) \\
	&= \sum_{m, \ell} \lambda_m \mu_\ell \qty[(P_m \circ Q_\ell)(P_m \circ Q_\ell)\T]_{ij},
	\end{align*}

	and therefore $K^{(1)} \circ K^{(2)} = \sum_{m, \ell} \lambda_m \mu_\ell \qty[(P_m \circ Q_\ell)(P_m \circ Q_\ell)\T]$, which is a nonnegative combination of positive semi-definite matrices and therefore positive semi-definite. 
\end{enumerate}

\end{proof}
\end{document}

### 2020-10-18 topics_statistical_theory/C1.tex
updates topics summary and example sheets for msm & topics

@@

    Furthermore, recall that the \emph{gamma function} is defined for $z \in (0, \infty)$ by
    \[
    \Gamma(z) = \int_0^\infty x^{z-1} e^{-x} \dd{x}
    \]
    and satisfies $\Gamma(n) = (n-1)!$ for all $n \in \NN$. 
\end{recap}
\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By Markov's inequality, we have for all $t \in \RR$ that
        \[
        \PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tX} \EE(e^{tX}) \leq e^{-tx + t^2\sigma^2/2}.  
        \]
        Since the LHS is independent of $t$, we can take the infimum over $t$ on the RHS and obtain
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$.

@@

\item Note that $x \mapsto e^{-tx}$ is convex for every $t \in \RR$, so $\EE(e^{-tX}) \geq e^{-t \EE(X)} \geq 1$ by Jensen's inequality.
    \end{enumerate}
\end{proof}

>>>>

	Furthermore, recall that the \emph{gamma function} is defined for $z \in (0, \infty)$ by
	\[
	\Gamma(z) = \int_0^\infty x^{z-1} e^{-x} \dd{x}
	\]
	and satisfies $\Gamma(n) = (n-1)!$ for all $n \in \NN$.

	Finally, recall the following inequality: for all $a, b \in \RR$ and $p \geq 1$
	\[
	(a+b)^p \leq 2^{p-1} (a^p + b^p). 
	\]
	This follows from the convexity of the function $x \mapsto x^p$. 
\end{recap}

@@

\begin{proof}
    \begin{enumerate}[(a)]
        \item We first consider $\PP(X \geq x)$. 
        By Markov's inequality, we have for all $t \in \RR$ that
        \[
        \PP(X \geq x) = \PP(e^{tX} \geq e^{tx}) \leq e^{-tX} \EE(e^{tX}) \leq e^{-tx + t^2\sigma^2/2}.  
        \]
        Since the LHS is independent of $t$, we can take the infimum over $t$ on the RHS and obtain
        \[
        \PP(X \geq x) \leq \inf_{t \in \RR} e^{-tx + t^2\sigma^2/2} = e^{-x^2/(2\sigma^2)}, 
        \]
        since the infimum of $t^2\sigma^2/2 - tx$ is attained at $t = x/\sigma^2$.

@@

\item Note that $x \mapsto e^{-tx}$ is convex for every $t \in \RR$, so $\EE(e^{-tX}) \geq e^{-t \EE(X)} = e^{0} = 1$ by Jensen's inequality. Let $X'$ denote an independent copy of $X$: then $X - X'$ has a symmetric distribution, so all its odd moments vanish. Therefore we find
\begin{align*}
	\EE[e^{tX}] &\leq \EE[e^{-tX'}] \EE[e^{tX}] = \EE[e^{t(X - X')}] = \EE \sum_{q=0}^\infty \qty[\frac{t^{2q} (X - X')^{2q}}{(2q)!}] \\
	&= \sum_{q=0}^\infty \frac{t^{2q} \EE[(X - X')^{2q}]}{(2q)!} \leq \sum_{q=0}^\infty \frac{2^{2q-1} t^{2q} \qty(\EE[X^{2q}] + \EE[(X')^{2q}])}{(2q)!} \\
	&\leq \sum_{q=0}^\infty  \frac{2^{2q - 1}t^{2q} 2 q! C^{2q}}{(2q)!} =\sum_{q=0}^\infty  \frac{(2tC)^{2q}q!}{(2q)!} = \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (q+j)} \\
	&\leq \sum_{q=0}^\infty \frac{(2tC)^{2q}}{\prod_{j=1}^q (2j)} = \sum_{q=1}^\infty \frac{(2t^2C^2)^q}{q!} = e^{2t^2C^2}.
\end{align*}
This shows that $X$ is sub-Gaussian with parameter $4C^2$. 
    \end{enumerate}
\end{proof}

Note that the proposition is not an ``if and only if''-type theorem: suppose we start with a sub-Gaussian variable $X$ with parameter $\sigma^2$. Then by (b), we have $\EE[X^{2q}] \leq q!(2\sigma)^{2q}$, and (c) then implies that $X$ is sub-Gaussian with parameter $16\sigma^2$.

\begin{theorem}[Hoeffding's inequality]
	Let $X_1, \dotsc, X_n$ be independent sub-Gaussian random variables, with $X_i$ having parameter $\sigma_i^2$. Then $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2}$. In particular, we have
	\[
	\max\qty{\PP(\bar X \geq x), \PP(\bar X \leq -x)} \leq e^{-nx^2/(2\overline{\sigma^2})}.
	\]
\end{theorem}

\begin{proof}
	For $t \in \RR$, we have
	\[
	\EE[e^{t\bar X}] = \EE[e^{(t/n)\sum_i X_i}] = \prod_{i=1}^n \EE[e^{(t/n)X_i}] \leq \prod_{i=1}^n e^{t^2\sigma_i^2/(2n^2)} = e^{t^2\overline{\sigma^2}/(2n)},
	\]
	which shows $\bar X$ is sub-Gaussian with parameter $\overline{\sigma^2}/n$. Applying part (a) of the previous proposition shows the second result. 
\end{proof}

\begin{remark}
	A direct consequence of Hoeffding's inequality is that
	\[
	\PP(\abs{\bar X} \geq x) \leq 2 e^{-nx^2/(2\overline{\sigma^2})}.
	\]
	The inequality is often stated in this weaker way. 
\end{remark}

\begin{lemma}[Hoeffding's lemma]
	Let $X$ be a random variable with $\EE X = 0$ that satisfies $a \leq X \leq b$. Then $X$ is sub-Gaussian with parameter $(b-a)^2/4$. 
\end{lemma}

\begin{proof}
	See Example Sheet 1, question 2. 
\end{proof}

### 2020-10-18 topics_statistical_theory/topics_ES1.tex
updates topics summary and example sheets for msm & topics

\begin{solution}
    First we compute the distribution function of $X_{(j)}$. If $X_{(j)} \leq x$, then at least $j$ of the $X_i$ must be $\leq x$.

For $k = j, j+1, \dotsc, n$, there are $\binom{n}{k}$ ways to choose $k$ of the $X_i$'s that must be $\leq j$. Fix $x \in [0, 1]$, then we have
\begin{align*}
    \PP(X_{(j)} \leq x) &= \sum_{k=j}^n \binom{n}{k} \PP(X_1 \leq x, \dotsc, X_k \leq x, X_{k+1} > x, \dotsc, X_n > x) \\
    &= \sum_{k=j}^n \binom{n}{k} (F(x))^k \cdot (1 - F(x))^{n-k} \\
    &=  \sum_{k=j}^n \binom{n}{k} x^k (1-x)^{n-k}.
\end{align*}

Now, the density function of $X_{(k)}$ is given by the derivative of $F$. This is a hefty calculation, see for example \href{https://math.stackexchange.com/a/1179378/696417}{this StackExchange link}, but it turns out to be
\[
\dv{x} F_{(j)}(x) = f_{(j)}(x) = \frac{n!}{(j-1)!(n-j)!} x^{j-1}(1-x)^{n-j},
\]
which is also the density function of the $\Rm{Beta}(j, n-j+1)$ distribution.

@@

\begin{question}
    Let $X_1, \dotsc, X_n$ be independent with distribution $P$ on a measurable space $(\Cal X, \Cal A)$, and let $\hat P_n$ be the empirical measure of $X_1, \dotsc, X_n$; thus $\hat P_n(A) = n^{-1} \sum_{i=1}^n \ind_{X_i \in A}$. Show that, for all $\eps > 0$ and $A \in \Cal A$, we have
    \[
    \PP\qty(\abs{\hat P_n(A) - P(A)} > \eps) \leq 2 e^{-2n\eps^2}. 
    \]
\end{question}

\begin{proof}
    Define a new distribution $Y = \ind_{X \notin A}$. Its distribution function is given by
    \[
    F_Y(y) = \begin{cases}
        0 &\text{$y < 0$}; \\
        P(A) &\text{$y \in [0, 1)$}; \\
        1 &y \geq 1. 
    \end{cases}, \quad F_Y(0) = P(A). 
    \]
    The empirical distribution function of $Y_1, \dotsc, Y_n \iid Y$ is given by
    \[
    \hat F_n(y) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq y},
    \]
    and thus for $y = 0$ we have
    \[
    \hat F_n(0) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq 0} = \frac1n \sum_{i=1}^n \ind_{X_i \in A} = \hat P_n(A).
    \]

>>>>

\begin{solution}
%    First we compute the distribution function of $U_{(j)}$. If $U_{(j)} \leq x$, then at least $j$ of the $U_{i}$ must be $\leq x$. 
%
%For $k = j, j+1, \dotsc, n$, there are $\binom{n}{k}$ ways to choose $k$ of the $U_{i}$'s that must be $\leq j$. Fix $x \in [0, 1]$, then we have
%\begin{align*}
%    \PP(U_{(j)} \leq x) &= \sum_{k=j}^n \binom{n}{k} \PP(X_1 \leq x, \dotsc, X_k \leq x, X_{k+1} > x, \dotsc, X_n > x) \\
%    &= \sum_{k=j}^n \binom{n}{k} (F(x))^k \cdot (1 - F(x))^{n-k} \\
%    &=  \sum_{k=j}^n \binom{n}{k} x^k (1-x)^{n-k}.
%\end{align*}
%
%Now, the density function of $X_{(k)}$ is given by the derivative of $F$. This is a hefty calculation, see for example \href{https://math.stackexchange.com/a/1179378/696417}{this StackExchange link}, but it turns out to be
%\[
%\dv{x} F_{(j)}(x) = f_{(j)}(x) = \frac{n!}{(j-1)!(n-j)!} x^{j-1}(1-x)^{n-j},
%\]
%which is also the density function of the $\Rm{Beta}(j, n-j+1)$ distribution.

We compute the density function of $U_{(j)}$ as follows: let $x \in (0, 1)$, then we know that 
\[
f_{(j)}(x) = \dv{x} F_{(j)}(x) = \lim_{h\to 0} \frac{F_{(j)}(x+h) - F_{(j)}(x)}{h} = \lim_{h \to 0} \frac{\PP(x < U_{(j)} \leq x + h)}{h}. 
\]
The probability $\PP(x < U_{(j)} \leq x + h)$ is the probability that exactly $j-1$ of the $U_{i}$ are less than $x$, and that at least one of the $U_{i}$ is in $(x, x+ h]$.

The probability that two or more of the $U_{i}$ lie in $(x, x+h]$ is $O(h^2)$ and therefore negligible, so we must compute the probability that exactly $j-1$ of the $U_{i}$ are smaller than $x$, one of the $U_{i}$ is in $(x, x+h]$, and the other $U_{i}$ are greater than $x+h$. This is easily seen to be
\begin{align*}
&\binom{n}{j-1} \PP(U \leq x)^{j-1} \cdot (n - j + 1) \PP(x < U \leq x + h) \cdot \PP(U > x + h)^{n-j} \\
&= \frac{n!}{(j-1)!(n-j+1)!} (n-j+1) x^{j-1} h (1 - x - h)^{n-j} \\
&= \frac{n!}{(j-1)!(n-j)!} x^{j-1} (1-x-h)^{n-j} h. 
\end{align*}
Therefore, we easily compute
\[
f_{(j)}(x) = \lim_{h \to 0} \frac{\frac{n!}{(j-1)!(n-j)!} x^{j-1} (1-x-h)^{n-j} h}{h} = \frac{n!}{(j-1)!(n-j)!} x^{j-1} (1-x)^{n-j}. 
\]
This is also the density function of a $\Beta(j, n-j+1)$ distribution.

@@

\begin{question}
    Let $X_1, \dotsc, X_n$ be independent with distribution $P$ on a measurable space $(\Cal X, \Cal A)$, and let $\hat P_n$ be the empirical measure of $X_1, \dotsc, X_n$; thus $\hat P_n(A) = n^{-1} \sum_{i=1}^n \ind_{U_{i} \in A}$. Show that, for all $\eps > 0$ and $A \in \Cal A$, we have
    \[
    \PP\qty(\abs{\hat P_n(A) - P(A)} > \eps) \leq 2 e^{-2n\eps^2}. 
    \]
\end{question}

\begin{proof}
    Define a new distribution $Y = \ind_{X \notin A}$. Its distribution function is given by
    \[
    F_Y(y) = \begin{cases}
        0 &\text{$y < 0$}; \\
        P(A) &\text{$y \in [0, 1)$}; \\
        1 &y \geq 1. 
    \end{cases}, \quad F_Y(0) = P(A). 
    \]
    The empirical distribution function of $Y_1, \dotsc, Y_n \iid Y$ is given by
    \[
    \hat F_n(y) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq y},
    \]
    and thus for $y = 0$ we have
    \[
    \hat F_n(0) = \frac1n \sum_{i=1}^n \ind_{Y_i \leq 0} = \frac1n \sum_{i=1}^n \ind_{U_{i} \in A} = \hat P_n(A).
    \]
